Evolutionary rescue by aneuploidy in tumors exposed to anti-cancer drugs

### 2025-02-03 ms.tex
revise fig S6-S9 captions and AppD

We distinguish between two scenarios for the proliferation time. 
In the first scenario, the tumor is small, $N\ll N_a^*\ll N_m^*$, and we have at most one lineage that rescues the cancer cell population.
As a result, the recurrence time is given by~\citep{avanzini2019cancer}
\begin{align}\label{meanproliferationtime}
\tau_a^r&\approx\tau_a+\frac{\log \left(p_mN\right)}{\Delta_m}.
\end{align}
The factor $p_m$ in the second term of \cref{meanproliferationtime} occurs because the lineage is conditioned to survive stochastic extinction and the time to reach $N$ is shorter compared to the scenario where the lineage is not conditioned to survive stochastic extinction~\citep{orr2014population,smith1974hitch}. % change 35

In the second scenario, the tumor is large, $N\gg N_m^*$, and the sensitive population produces a large number of rescue lineages in a short period of time.
As a result, the recurrence time is obtained by solving the following system of ordinary differential equations (ODEs),
\begin{equation}\label{detODE}
\begin{aligned}
\frac{ds}{dt}&=\Delta_ss,\\
\frac{da}{dt}&=\Delta_aa+u\lambda_ss,\\
\frac{dm}{dt}&=\Delta_mm+v\lambda_aa+v\lambda_ss.
\end{aligned}
\end{equation}
Solving the system of ODEs for initial condition $\left(s(0), a(0), m(0)\right)=\left(N,0,0\right)$ we obtain:
\begin{equation*} \label{m_t}
m(t) = \frac{N u v \lambda_a \lambda_s}{\Delta_a - \Delta_s}\left[ \frac{\e^{\Delta_m t} - \e^{\Delta_a t}}{\Delta_m - \Delta_a} - \frac{\e^{\Delta_m t} - \e^{\Delta_s t}}{\Delta_m - \Delta_s} \right] + N v \lambda_s \frac{\e^{\Delta_m t} - \e^{\Delta_s t}}{\Delta_m - \Delta_s}.
\end{equation*}
We obtain $\tau_a^r$ by solving the above equation for time $t=\tau_a^r$ at which the number of mutant cells reaches $N$, that is, we solve $m\left(\tau_a^r\right)=N$,
\begin{equation}\label{eq:t2det2024}
1 = \frac{u v \lambda_a \lambda_s}{\Delta_a - \Delta_s}\left[ \frac{\e^{\Delta_m \tau_a^r} - \e^{\Delta_a \tau_a^r}}{\Delta_m - \Delta_a} - \frac{\e^{\Delta_m \tau_a^r} - \e^{\Delta_s \tau_a^r}}{\Delta_m - \Delta_s} \right] + v \lambda_s \frac{\e^{\Delta_m \tau_a^r} - \e^{\Delta_s \tau_a^r}}{\Delta_m - \Delta_s}.
\end{equation}

\Cref{eq:t2det2024} is a transcendental equation which cannot be solved exactly but we can obtain an approximation by noting that the tumor size $N$ is large, and because the sensitive and aneuploid cells have negative growth rates, almost all of the growth needed to rebound to size $N$ will be by mutant cells. 
Mathematically, this corresponds to $\e^{\Delta_m \tau_a^r}$ being much larger than the other exponential terms,  $\e^{\Delta_s \tau_a^r}$ and $\e^{\Delta_a \tau_a^r}$. 
\Cref{eq:t2det2024} can then be rewritten as
\[
	1 \approx \frac{v \lambda_s e^{\Delta_m \tau_a^r}}{\Delta_m - \Delta_s} 
	\left(\frac{u \lambda_a}{\Delta_m - \Delta_a} + 1\right).
\]
Assuming that rate of aneuploidy is not so high that it overwhelms the aneuploids' growth disadvantage relative to the mutants ($u \lambda_a \ll \Delta_m - \Delta_a$), we can neglect the first term in parentheses above, yielding
\begin{equation}\label{eq:t2det}
\tau_a^r\approx\frac{1}{\Delta_m}\log{\left(\frac{\Delta_m-\Delta_s}{v\lambda_s}\right)}.
\end{equation}
We emphasize that in this case the population is large enough to produce mutants without needing to pass through aneuploidy ($N \gg N_m^*$), and so  the terms arising from the evolutionary trajectory \textit{sensitive} $\rightarrow$ \textit{aneuploid} $\rightarrow$ \textit{mutant} do not contribute to the above approximation. %change 9

@@

From  \cref{eqmNs} we obtain the distribution of the recurrence time conditional of evolutionary rescue: %change 12
\begin{align}\label{distribution}
f\left(t\right)&=\frac{d}{dt}\left[\frac{P\left(m_t\geq N\right)}{\presc}\right]=r'\left(t\right)\frac{\exp\left[-r\left(t\right)\right]}{\presc},
\end{align}
which we plot in \Cref{KMdistribution} and compare with simulations. We note that in the scenario $N\gg\abs{\Delta_s}\abs{\Delta_a}/uv\lambda_s\lambda_ap_m$ the distribution becomes the Dirac $\delta$-function~\citep{barton1989elements}.
%%%
\section{Bootstrapping}
For the mean times the $95\%$ confidence interval is obtained through bootstrapping in the following steps: (1) we simulate $T$ 100 times; (2) we sample with replacement which we store in $T'$; (3) for each element of this sample we obtain $\tau=\mathbb{E}\left[T'\right]$; (4) we repeat steps (2)-(3) 100 times to obtain $\tau$ and we select the upper and lower limits such that $95\%$ of the values of $\tau$ lie in the interval given by the bounds.

@@

\caption{Shown is the mean time for appearance of a resistance mutation that leads to evolutionary rescue with aneuploidy ($u>0$). Our inhomogeneous Poisson-process approximations (solid black lines: \cref{meantimet2}) are in agreement with simulation results (red markers with 95\% confidence intervals obtained with bootstrapping, see Appendix G). Dashed vertical blue line represents the threshold tumor size above which evolutionary rescue is very likely through aneuploidy \cref{eq:N_a} and the dashed vertical green line represents the threshold tumor size above which evolutionary rescue is very likely through direct mutation \cref{eq:N_m}. Solid lines represent the approximations \cref{eq:AsymptoticTimeRules} ($N<N_a^*$ blue line and $N>N_m^*$ green line). 

@@

Our inhomogeneous Poisson-process approximation (solid black line, \cref{meanproliferationtime}) is in agreement with simulation results (red markers with 95\% confidence intervals obtained with bootstrapping, see Appendix G) for small and intermediate values of $N$. Dashed vertical blue line represents the threshold tumor size above which evolutionary rescue is very likely through aneuploidy \cref{eq:N_a} and the dashed vertical green line represents the threshold tumor size above which evolutionary rescue is very likely through direct mutation \cref{eq:N_m}. Solid lines represents the approximations \cref{eq:AsymptoticTimeRulesRecurrence} ($N<N_a^*$ blue line and $N>N_m^*$ green line). The simulations converge to \cref{eq:t2det} (green line) for large values of $N\gg N_m^*$.  Parameters: $\lambda_s=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.}

@@

Our inhomogeneous Poisson-process approximation (solid black line, \cref{meanproliferationtime2}) is in agreement with simulation results (red markers with 95\% confidence intervals obtained with bootstrapping, see Appendix G) for small and intermediate values of $N$. Dashed vertical blue line represents the threshold tumor size above which evolutionary rescue is very likely through aneuploidy \cref{eq:N_a} and the dashed vertical green line represents the threshold tumor size above which evolutionary rescue is very likely through direct mutation \cref{eq:N_m}. Solid blue line represents the approximation \cref{meanproliferationtime2} with $\tau_a$ from \cref{eq:AsymptoticTimeRules} for $N<N_a^*$ and the solid green line  represents the approximation \cref{eq:t2det2} for $N>N_m^*$. The simulations converge to \cref{eq:t2det2} (green line) for large values of $N\gg N_m^*$.  Parameters: $\lambda_s=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}, M=10^7$.}

@@

\caption{Shown is the mean time for appearance of a resistance mutation that leads to evolutionary rescue with aneuploidy ($u>0$) when a fraction $f$ of cancer cells are aneuploid at the start of therapy. Black lines represent our inhomogeneous Poisson-process approximations (solid black line, \cref{meantimet2}; dashed black line \cref{meantimet2SGV}). Dashed black line is the  inhomogeneous Poisson-process approximation where a fraction $f$ of tumor is aneuploid at the onset of drug therapy which is in agreement with simulation results (red markers with 95\% confidence intervals obtained with bootstrapping, see Appendix G). Dashed vertical blue line represents the threshold tumor size above which evolutionary rescue is very likely through aneuploidy \cref{eq:N_a} and the dashed vertical green line represents the threshold tumor size above which evolutionary rescue is very likely through direct mutation \cref{eq:N_m}. Solid lines represents the approximations \cref{eq:AsymptoticTimeRules} ($N<N_a^*$ blue line and $N>N_m^*$ green line). 

>>>>

We distinguish between small, intermediate, or large tumors.
In small ($N\ll N_a^*\ll N_m^*$) and intermediate ($N_a^* \ll N \ll N_m^*$) tumors, we have at most one lineage that rescues the cancer cell population.
As a result, the recurrence time is given by~\citep{avanzini2019cancer}
\begin{align}\label{meanproliferationtime}
\tau_a^r&\approx\tau_a+\frac{\log \left(p_mN\right)}{\Delta_m}.
\end{align}
The factor $p_m$ in the second term of \cref{meanproliferationtime} occurs because the lineage is conditioned to survive stochastic extinction and the time to reach $N$ is shorter compared to the scenario where the lineage is not conditioned to survive stochastic extinction~\citep{orr2014population,smith1974hitch}. % change 35 
In both scenarios, plugging \cref{meantimet2} for $\tau_a$ in \cref{meanproliferationtime} provides a good approximation for the recurrence time (black line in \Cref{ProliferationTimeLarge}).
For small tumors, we have a simpler approximation by plugging \cref{limitapprox4} for $\tau_a$ to get $\tau_a^r\approx1/|\Delta_a| + \log{\left(p_mN\right)}/\Delta_m$ (blue line in \Cref{ProliferationTimeLarge}).

In a large tumor, $N_m^* \ll N$, the sensitive population produces a large number of rescue lineages in a short period of time.
As a result, the recurrence time is obtained by solving the following system of ordinary differential equations (ODEs),
\begin{equation}\label{detODE}
\begin{aligned}
\frac{ds}{dt}&=\Delta_ss,\\
\frac{da}{dt}&=\Delta_aa+u\lambda_ss,\\
\frac{dm}{dt}&=\Delta_mm+v\lambda_aa+v\lambda_ss.
\end{aligned}
\end{equation}
Solving the system of ODEs for initial condition $\left(s(0), a(0), m(0)\right)=\left(N,0,0\right)$ we obtain:
\begin{equation*} \label{m_t}
m(t) = \frac{N u v \lambda_a \lambda_s}{\Delta_a - \Delta_s}\left[ \frac{\e^{\Delta_m t} - \e^{\Delta_a t}}{\Delta_m - \Delta_a} - \frac{\e^{\Delta_m t} - \e^{\Delta_s t}}{\Delta_m - \Delta_s} \right] + N v \lambda_s \frac{\e^{\Delta_m t} - \e^{\Delta_s t}}{\Delta_m - \Delta_s}.
\end{equation*}
We obtain $\tau_a^r$ by solving the above equation for time $t=\tau_a^r$ at which the number of mutant cells reaches $N$, that is, we solve $m\left(\tau_a^r\right)=N$,
\begin{equation}\label{eq:t2det2024}
1 = \frac{u v \lambda_a \lambda_s}{\Delta_a - \Delta_s}\left[ \frac{\e^{\Delta_m \tau_a^r} - \e^{\Delta_a \tau_a^r}}{\Delta_m - \Delta_a} - \frac{\e^{\Delta_m \tau_a^r} - \e^{\Delta_s \tau_a^r}}{\Delta_m - \Delta_s} \right] + v \lambda_s \frac{\e^{\Delta_m \tau_a^r} - \e^{\Delta_s \tau_a^r}}{\Delta_m - \Delta_s}.
\end{equation}

\Cref{eq:t2det2024} is a transcendental equation which cannot be solved exactly but we can obtain an approximation by noting that the tumor size $N$ is large, and because the sensitive and aneuploid cells have negative growth rates, almost all of the growth needed to rebound to size $N$ will be by mutant cells. 
Mathematically, this corresponds to $\e^{\Delta_m \tau_a^r}$ being much larger than the other exponential terms,  $\e^{\Delta_s \tau_a^r}$ and $\e^{\Delta_a \tau_a^r}$. 
\Cref{eq:t2det2024} can then be rewritten as
\[
	1 \approx \frac{v \lambda_s e^{\Delta_m \tau_a^r}}{\Delta_m - \Delta_s} 
	\left(\frac{u \lambda_a}{\Delta_m - \Delta_a} + 1\right).
\]
Assuming that rate of aneuploidy is not so high that it overwhelms the aneuploids' growth disadvantage relative to the mutants ($u \lambda_a \ll \Delta_m - \Delta_a$), we can neglect the first term in parentheses above, yielding (green line in \Cref{ProliferationTimeLarge})
\begin{equation}\label{eq:t2det}
\tau_a^r\approx\frac{1}{\Delta_m}\log{\left(\frac{\Delta_m-\Delta_s}{v\lambda_s}\right)}.
\end{equation}
We emphasize that in this case the population is large enough to produce mutants without needing to pass through aneuploidy ($N \gg N_m^*$), and so  the terms arising from the evolutionary trajectory \textit{sensitive} $\rightarrow$ \textit{aneuploid} $\rightarrow$ \textit{mutant} do not contribute to the above approximation. %change 9

@@

From  \cref{eqmNs} we obtain the distribution of the recurrence time conditional of evolutionary rescue: %change 12
\begin{align}\label{distribution}
f\left(t\right)&=\frac{d}{dt}\left[\frac{P\left(m_t\geq N\right)}{\presc}\right]=r'\left(t\right)\frac{\exp\left[-r\left(t\right)\right]}{\presc},
\end{align}
which we plot in \Cref{KMdistribution} and compare with simulations. We note that in the scenario $N\gg\abs{\Delta_s}\abs{\Delta_a}/uv\lambda_s\lambda_ap_m$ the distribution becomes the Dirac $\delta$-function~\citep{barton1989elements}.
%%%
\section{Bootstrap}
For the mean times the $95\%$ confidence interval is obtained through bootstrapping in the following steps: (1) we simulate $T$ 100 times; (2) we sample with replacement which we store in $T'$; (3) for each element of this sample we obtain $\tau=\mathbb{E}\left[T'\right]$; (4) we repeat steps (2)-(3) 100 times to obtain $\tau$ and we select the upper and lower limits such that $95\%$ of the values of $\tau$ lie in the interval given by the bounds.

@@

\caption{The mean time for appearance of a resistance mutation that leads to evolutionary rescue with aneuploidy ($u>0$). 
Solid lines show our approximations (\cref{eq:AsymptoticTimeRules}) compared to red markers that show mean of simulations results (with error bars for 95\% confidence intervalues obtained with bootstrap, see Appendix G).
Blue dashed line, $N_a^*$.
Green dashed line, $N_m^*$.

@@

Solid lines show our approximations (green: \cref{eq:t2det}, blue: \cref{meanproliferationtime} with \cref{limitapprox4} for $\tau_a$; black: \cref{meanproliferationtime} with \cref{meantimet2} for $\tau_a$) compared to red markers that show mean of simulations results (with error bars for 95\% confidence intervals obtained with bootstrap, see Appendix G).
Blue dashed line, $N_a^*$.
Green dashed line, $N_m^*$.
Parameters: $\lambda_s=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.}

@@

Solid lines show our approximations (green: \cref{eq:t2det2} for $N>N_m^*$, blue:  \cref{meanproliferationtime2} with $\tau_a$ from \cref{eq:AsymptoticTimeRules} for $N<N_a^*$; black: \cref{meanproliferationtime2}) compared to red markers that show mean of simulations results (with error bars for 95\% confidence intervals obtained with bootstrap, see Appendix G).
Blue dashed line, $N_a^*$.
Green dashed line, $N_m^*$.
Parameters: $\lambda_s=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}, M=10^7$.}

@@

\caption{The mean time for appearance of a resistance mutation that leads to evolutionary rescue with aneuploidy ($u>0$) when a fraction $f$ of cancer cells are aneuploid at the start of drug treatment. 
Lines show our approximations (solid green: \cref{eq:AsymptoticTimeRules} for $N>N_m^*$; solid blue:  \cref{eq:AsymptoticTimeRules} for $N<N_a^*$; solid black: \cref{meantimet2}; dashed black: \cref{meantimet2SGV}) compared to red markers that show mean of simulations results (with error bars for 95\% confidence intervals obtained with bootstrap, see Appendix G).
Blue dashed line, $N_a^*$.
Green dashed line, $N_m^*$.
